---
title: "National and Seasonal Effects on COVID Deaths "
author: "Oscar Rivera"
date: "3/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(lubridate)
library(gplots)
library(car)
library(MASS)
library(tidyverse)
```

```{r include=FALSE}
covid <- read.csv("https://covid19.who.int/WHO-COVID-19-global-data.csv")
```

# Introduction

 When the COVID-19 virus made its presence in North America, it was very quick 
to make a negative impact on the nations in this continent, with those nations 
being the United States, Canada, and Mexico. Having settled in these nations, 
the virus brought a significant number of infections and deaths to those living in these 
nations within first two years of its presence. Throughout these two years, there were certain 
time periods in which the number of deaths due to COVID rose 
up drastically. Furthermore, these number of deaths turned out to be much higher 
for some nations. Since the virus became a threat to the world, The World Health 
Organization (WHO) has been monitoring the number of deaths and cases per for 
various countries around the globe. The data set which WHO uses to keep track 
of these number can observed through the [weekly WHO COVID-19 update](https://www.who.int/emergencies/diseases/novel-coronavirus-2019/situation-reports).
For this project, we will focus on determining 
whether the nations in North America and the seasons of the year have major 
effects on the number of deaths caused by COVID-19 during the pandemic. 

# Background

 The data set that we will be working with to answer the following question is 
WHO COVID-19 data set that was mentioned in the introduction. Through this data 
set, WHO has managed to keep track of the daily and cumulative number of COVID 
cases and deaths between January 20202 to present day as well as for all nations 
across the globe. 

```{r echo=FALSE}
#Shortening data to US, CA, and MX
US <- which(covid$Country_code=="US")
CA <- which(covid$Country_code=="CA")
MX <- which(covid$Country_code=="MX")

covid_NA <- covid[c(US,CA,MX),]
```

Since our question of interest is focused on countries in North America, we 
minimized the observations to those involving the United States, Canada, and 
Mexico. Around March 2020, the number of COVID cases and deaths became more 
recognized in North America in the sense that the nations in this continent
began taking serious measures to control for these number. 
With this being said, I have restricted this data set to dates between March 1, 
2020 to February 28, 2022. In addition, applying this restriction on the data 
set give more flexibility to use appropriate methods in my analysis, which will 
be discussed later. 

```{r echo=FALSE}
#Creating Month and Year variables
covid_NA$Month <- month(covid_NA$Date_reported)
covid_NA$Year <- year(covid_NA$Date_reported)

Jan2020 <- which(covid_NA$Month==1 & covid_NA$Year==2020)
Feb2020 <- which(covid_NA$Month==2 & covid_NA$Year==2020)

covid_NA <- covid_NA[-c(Jan2020,Feb2020),]

#Creating Season variable 
covid_NA$Season <- c()
n_T <- nrow(covid_NA)

for (i in 1:n_T) {
  if (covid_NA$Month[i] == 1 | covid_NA$Month[i] == 2 | covid_NA$Month[i] == 12){
    covid_NA$Season[i] <- "Winter"
  }
  if (covid_NA$Month[i] == 3 | covid_NA$Month[i] == 4 | covid_NA$Month[i] == 5){
    covid_NA$Season[i] <- "Spring"
  }
  if (covid_NA$Month[i] == 6 | covid_NA$Month[i] == 7 | covid_NA$Month[i] == 8){
    covid_NA$Season[i] <- "Summer"
  }
  if (covid_NA$Month[i] == 9 | covid_NA$Month[i] == 10 | covid_NA$Month[i] == 11){
    covid_NA$Season[i] <- "Fall"
  }
}
covid_NA$Season <- as.factor(covid_NA$Season)
covid_NA$Country_code <- as.factor(covid_NA$Country_code)

covid_NA <- covid_NA[,-3]

# Removing all dates of March 2022 and beyond
March_2022 <- which(covid_NA$Month >= 3 & covid_NA$Year == 2022)
covid_NA <- covid_NA[-March_2022,]

rownames(covid_NA) <- 1:nrow(covid_NA)

#Creating Case Fatality Ratio Variable
covid_NA$New_CFR <- c()
for (i in 1:nrow(covid_NA)) {
  if(covid_NA$New_cases[i] == 0 ){
    covid_NA$New_CFR[i] = 0 
  }
  else{
    covid_NA$New_CFR[i] <- covid_NA$New_deaths[i]/covid_NA$New_cases[i]
  }
}
#View(covid_NA)
```

To create a variable for season, we used the month number 
of the dates listed in the data. All observations with month numbers 1, 2, or 
12 are assigned. Winter for season, all dates with month numbers 3, 4, or 5 are 
assigned Spring for season, all dates with month numbers 6, 7, or 8 are assigned 
Summer for season, and all dates with month numbers 9, 10, and 11 are assigned 
Fall for season. Using the number of deaths and the number of cases per day,
we constructed a variable for case-fatality rate (CFR) per day, which is 
defined as
\[
   CFR = \dfrac{\text{Number of COVID deaths per day}}{\text{Number of COVID deaths per day}}
\]
Viewing the WHO data set, one will notice that the for some dates, the number of
cases are zero. For those dates, we set the CFR equal to zero. 

# Exploratory Data Analysis

 Restricting the WHO data to the United States, Canada, and Mexico and to the 
dates March 1 2020 - February 28 2022, the modified
data set contains `r nrow(covid_NA)` total observations.  

## Time Series Plots

To study the daily number of deaths, cases, and case-fatality rates due to COVID 
within a span of two years, I created a times series plot of these numbers for 
all three nations. 

```{r echo=FALSE}
US_NA <- which(covid_NA$Country_code=="US")
CA_NA <- which(covid_NA$Country_code=="CA")
MX_NA <- which(covid_NA$Country_code=="MX")

t <- 1:length(covid_NA$New_deaths[US_NA])

par(mfrow=c(2,2))
plot(t,covid_NA$New_deaths[US_NA],type='l', main="Number of Deaths due to 
     COVID-19", xlab="Time: March 1, 2020 - February 28, 2022", ylab="Number of Deaths")
lines(t,covid_NA$New_deaths[CA_NA],type='l',col=3)
lines(t,covid_NA$New_deaths[MX_NA],type='l',col=4)
legend("topleft", legend = c("US","CA","MX"), lty=c(1,1,1), col=c(1,3,4),
       cex=0.5)

plot(t,covid_NA$New_cases[US_NA],type='l', main="Number of Cases due to 
     COVID-19", xlab="Time: March 1, 2020 - February 28, 2022", ylab="Number of Deaths")
lines(t,covid_NA$New_cases[CA_NA],type='l',col=3)
lines(t,covid_NA$New_cases[MX_NA],type='l',col=4)
legend("topleft", legend = c("US","CA","MX"), lty=c(1,1,1), col=c(1,3,4),
       cex=0.5)
#par(mfrow=c(1,1))

plot(t,covid_NA$New_CFR[US_NA],type='l', main="Case Fatality Rate due to 
     COVID-19", xlab="Time: March 1, 2020 - February 28, 2022", ylab="Case
     Fatality Rate",ylim=c(0,.5))
lines(t,covid_NA$New_CFR[CA_NA],type='l',col=3)
lines(t,covid_NA$New_CFR[MX_NA],type='l',col=4)
legend("topright", legend = c("US","CA","MX"), lty=c(1,1,1), col=c(1,3,4),
       cex=0.5)
par(mfrow=c(1,1))
```

Looking at the time series plot of the COVID deaths for the three nations 
(topleft), all three countries appear to show similar trends. In all time 
periods, Canada tends to have the lowest number of deaths despite being the 
largest country from the three. Although it seems that COVID deaths are higher 
for the US, one can see that these numbers for all three nations tend to 
fluctuate roughly around the same time of the year. This may be an indication 
that the seasons of each year may play a role on the number of deaths rising up 
so rapidly. 
In the time series plot of the COVID cases for the three countries (topright), 
the US tends to have the highest number of cases. Further, it shows that the US
had quite some time periods where the cases got significantly high. Near the 
ending of this plot, one can observe that the cases for all three nation 
increased around the same time period. 
In the time series plot of the CFR for the three countries (bottomleft), Mexico
appears to have the highest CFR. Note that Canada has a very high CFR value 
for one of the dates. The reason for the CFR being that high is because on that 
day, there two COVID cases and one COVID, leading to $CFR=1/2$. The date corresponding
to this is March 11, 2020, which is the day Canada reported it's first COVID 
death.

## Sample Averages

  To study the average numbers number of deaths caused by COVID in it's two-year 
present, we provided these average numbers by country and season in the table 
below. 

```{r echo=FALSE}
#aggregate(New_deaths~Country_code+Season,data=covid_NA,mean)
aggregate(list(Deaths=covid_NA$New_deaths,Cases=covid_NA$New_cases,
               CFR = covid_NA$New_CFR),
          by=list(Country=covid_NA$Country_code,Season=covid_NA$Season), mean)
```

Viewing this table, one can observe that the average number of deaths and 
average number of case due to COVID is highest for the US in all seasons. Notice 
that in spring seasons, the average cases for Canada are higher than that
of Mexico; however, the average cases are a lot lower for Canada during the 
summer time. The highest average case fatality rate goes to Mexico during the 
spring season. One interesting observation 
to notice is that the average number of deaths and the average number 
of cases caused by COVID tend to be highest for all three nation during the 
winter time. 

## Five-Number Summary

We will now explore the five-number summary of COVID deaths, cases, and case-
mortality rates. Recall that the five-number summary consist 5 measures,
the minimum, 1st quartile (Q1), mediam (Q2), 3rd quartile (Q3), and the maximum. 

The following table displays the five-number summary of COVID Deaths by country
and season:

```{r echo=FALSE, message=FALSE, warning=FALSE}
covid_NA %>% 
  group_by(Country_code,Season) %>% 
  summarise(min = fivenum(New_deaths)[1],
            Q1 = fivenum(New_deaths)[2],
            median = fivenum(New_deaths)[3],
            Q3 = fivenum(New_deaths)[4],
            max = fivenum(New_deaths)[5])
```

The table shows that Spring is the only season where the minimum number 
of COVID deaths for the three countries is 0. This is because most deaths 
did not occur in these countries the begin of Spring in 2020. Notice that the
median number of deaths for Mexico and the US are not too apart from each other.
The spring time brought the maximum number of deaths for Canada, whereas the 
winter time brought the number of maximum number of deaths for the US and 
Mexico. 

The next following table shows the five-number summary of COVID Cases by country
and season:

```{r echo=FALSE, message=FALSE, warning=FALSE}
covid_NA %>% 
  group_by(Country_code,Season) %>% 
  summarise(min = fivenum(New_cases)[1],
            Q1 = fivenum(New_cases)[2],
            median = fivenum(New_cases)[3],
            Q3 = fivenum(New_cases)[4],
            max = fivenum(New_cases)[5])
```

Notice that the minimum number of cases for Canada in all seasons is 0. This 
indicates that Canada had days in all seasons where no cases were reported. 
Neither the US nor Mexico had 0 cases in any seasons during this two-year 
timeline. The median number of cases appears to be highest for the US. 
The table also shows that the median and maximum number of cases in all three 
nations were highest during the winter time.

From both tables, it appears that both nations and seasons may have some role
in the number of deaths and cases. 

```{r message=FALSE, warning=FALSE, include=FALSE}
covid_NA %>% 
  group_by(Country_code,Season) %>% 
  summarise(min = round(fivenum(New_CFR)[1],4),
            Q1 = round(fivenum(New_CFR)[2],4),
            median = round(fivenum(New_CFR)[3],4),
            Q3 = round(fivenum(New_CFR)[4],4),
            max = round(fivenum(New_CFR)[5],4))
```

## Interaction Plot of Average COVID Deaths by Country and Season

To examine the effects these seasons and nations have on the death number 
created an interaction plot of the average COVID deaths numbers by season and
nationality. 

```{r echo=FALSE}
interaction.plot(covid_NA$Country_code,covid_NA$Season,covid_NA$New_deaths,
          xlab="Country", ylab="Average Number of Deaths by COVID",type="l", 
          lty=1, lwd=2, legend=T,col=c(2,3,4,5),trace.label="Season" )
```

The differences in the heights of the four curves reflects the seasonal effects 
on deaths, and the deviation from horizontal reflects the effects of nationality 
on deaths. The plot indicates that there may be effects from nationality and 
season present in the number deaths caused by COVID. 
The curves for Winter, Fall, and Spring appear to have similar behaviors look 
to be nearly parallel. However, one may notice that the curves for for Fall, 
Spring, and Summer appear to intersect on the left portion of the 
plot. Also, the curve for Summer intersects with both the curves of Fall and 
Spring in between Mexico and US. This observation points out that there may be
an Nationality-Seasonal interaction effect on the number of deaths. 

# Model Proposal

To determine whether the nations in North America and the seasons in a calendar 
year have major effects on the number of deaths caused by COVID-19, we will be 
using a balanced Two-Way ANOVA model. A balanced Two-Way ANOVA model in 
factor-effect is written in the following form:
\[
   Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \epsilon_{ijk}, 
   \ i = 1, \dots, a; \ \
   j = 1, \ldots, b; \ \ k = 1, \ldots, n_{ij}.
\] 
where, $\alpha_i$ is the fixed main effect of factor A at the $i$th level, 
$\beta_j$ is the fixed main effect of factor B at the $j$th level, and 
$(\alpha \beta)_{ij}$ is the interaction effect when factor A is at the $i$th 
level and factor B is at the $j$th level. For this model to be valid, it must 
have the following constraints:
\[
  \sum_{i=1}^a \alpha_i = 0, \ \ \ \ \sum_{j=1}^b \beta_j = 0, 
\]
\[
  \sum_{i=1}^a (\alpha \beta)_{ij} = 0 \ \ \ \text{and} \ \ \ 
  \sum_{j=1}^b (\alpha \beta)_{ij} = 0
\]
and it must meet the assumption that $\varepsilon_{ijk}$ are i.i.d. 
(independently and identically distributed) as $N(0,\sigma^2)$. 

A Two-Way Anova model is balance if the number of observation by two categories
are equal. The following table displays the number of observations by season and 
nationality.

```{r echo=FALSE}
table(covid_NA$Country_code,covid_NA$Season)
```

Notice that these number are slightly different. This is because both Spring 
and Summer consist of two months with 31 days and one month with 30 days. The 
number of observations are smallest for Winter since it contains the month 
of February, which has the shortest number of days compared to all other months. 
Since the observations are large and do not differ much in size, we can say the 
a Two-Way ANOVA model applied to this data would be approximately balanced.

I believe that this model is appropriate for my task on studying the effects
nationality and season have on the COVID death numbers since Two-Way ANOVA 
models give flexibility when the studying a quantitative variable using two 
categorical variables. Moreover, Two-Way ANOVA provides the ability to test
whether main effects of each of the two categorical variables are present in 
the quantitative variable. Using a Two-Way ANOVA model for my question of 
interest, we will have the following representations:

* $Y_{ijk}$ is the cube-rooted number of deaths caused by COVID in the $i$th
  nation during the $j$th season. 
* $\alpha_i$ is the main factor effect of nationality at the $i$th nation 
  for $i = 1,\ldots,a=3$ (three nations),
* $\beta_j$ is the main effect of seasonality at the $j$th season for 
  $j = 1, \ldots, b=4$ (four seasons), 
* $(\alpha \beta)_{ij}$ is the interaction effect when of the $i$th nation 
  and $j$th season.

# Model Fitting

Fitting a Two-Way ANOVA model using the cube-rooted number of COVID deaths, 
nationality and Season, we obtain the following results:

```{r echo=FALSE}
anova_model <- aov(I(New_deaths)^(1/3)~Country_code*Season,data=covid_NA)
summary(anova_model)
```

Note that the Country_code, Season, and the interaction between Country_Code and 
season have corresponding $P$-values less than $2\times 10^{-16}$, which are very 
small $P$-values. This indicates that there is sufficient evidence to conclude 
that there are main effects from Nationality, main effects from Season, and 
interaction effects present in the cube-rooted number of deaths caused by COVID. 

# Model Diagnostics

I will now explore plots involving the residuals from the Two-Way ANOVA to 
see whether the model assumption that $\varepsilon_{ijk}$ are i.i.d as 
$N(0,\sigma^2)$. 

```{r echo=FALSE}
par(mfrow=c(2,2))
plot(anova_model,which=c(1,2))
hist(anova_model$residuals,main="Histogram of Residuals",
     ylab="Residuals")
```

Viewing the Residuals vs Fitted plot, one can see that variability among the 
residuals are not necessarily the same. Furthermore, the red-line in this plot
deviates slightly from the dash-line. This indicates that the error terms 
do not necessarily have constance variance and mean 0. Looking at the Normal 
Q-Q plot, it appears that most of the points are on the 
gray dashed line. However, one can see that the points on the left deviate from
the dashed-line. This indications that the distribution of the residuals is 
slightly left-skewed. The histogram of the residuals shows that the distribution
of the residual is slight left-skewed and looks to be almost symmetric. 
Since the Two-Way ANOVA model that was fitted is approximately balanced, the 
results obtained from the model fitting are pretty robust against the normality
assumption and unequal variance. 

```{r echo=FALSE}
acf(anova_model$residuals,main="ACF plot of Residuals")
```

Since all bars shown in the plot stretch above the upper blue dashed-line, this
suggests that the error terms are not necessarily dependent. One reason why 
this is the case in my situation is because the data I'm working with are 
time observations, which are not randomized. 

# Discussion and Conclusion 

Although it appears that model assumptions for the Two-Way ANOVA model that I 
fit weren't necessarily, the results obtained from the fitting process are robust
again most of these assumptions since the model is approximately balanced. Since
the data im working with is time series data and observational data, thus making
it difficult to correct for non-independence of the residuals. Although, the 
model give supporting evidence that season has effects in the number of COVID deaths
per day, there maybe be other factors along with season that may cause the 
number of deaths to fluctuate. In the future, I would like to learn whether 
there are other variables play a role in these numbers. Indeed, each season of 
the year bring various challenges to these nations, such as cooler and warmer 
temperatures and flu season, that may potentially cause the COVID death numbers 
for these three nations to go up. 
# References

Kutner, M. H., Nachtsheim, C. J., Neter, J., &amp; Li, W. (2005). 
Applied Linear Statistical Models / Michael H. Kutner ... et al.. McGraw-Hill Irwin.

“Coronavirus Disease (COVID-19) Situation Reports.” World Health Organization, World Health Organization, https://www.who.int/emergencies/diseases/novel-coronavirus-2019/situation-reports. 

# Appendix - R Code used {-}

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```

# Acknowledgements

* Used STA 232A notes for reference. 
